---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Yogi Patnaik"
date: "September 11, 2026"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---



```{r}
library(tidyverse)
library(tidycensus)

```


In this analysis, I look at median gross rent across counties in California and compare how reliable the estimates are. I use the margin of error to identify which counties have more or less uncertainty in their rent estimates.

---





```r
vars <- load_variables(2023, "acs5")
View(vars)
```


```{r}
my_state <- "CA"       
my_variable <- "B25064_001"    

county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

```{r}
dim(county_data)
head(county_data)
```


The state of California has 58 counties, and the row count was also 58.




---


```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
  
```



```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```




Sierra County had the largest margin of error relative to the estimate, with a margin of error percentage of 59.7%.

---

 Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```




**[How many counties fall into each category?]**
From the total 58 counties in California, 39 counties were classified as high confidence, 13 as moderate confidence, and 6 as low confidence.
---

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```



```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```




** [What is the average population of high-confidence counties versus low-confidence
counties? Is there a pattern?]**

The average population of high confidence counties is  989,117.82, and the average population of low confidence is  11,691.33. The pattern shows that less reliable estimates tend to occur in counties that have smaller populations.

---



 We know the least about the smaller counties in California. Sierra County had a highest margin of error relative to its median gross rent estimate at 59.7%, and other smaller counties, like Trinity County, Plumas County, Mariposa County, and Mono County, had very high margins of error. These six counties, which were classified as low confidence had an average county population of 11,691. 

 Since the ACS only uses a sample of households rather than surveying every household in the country, there is less samples and therefore more uncertainty and larger margins of error when collecting data. The smaller counties having less reliable median gross rent estimates is explained by this.

 This uncertainty plays a part in a situation where housing funding may be allocated based on an equally certain approach from a state agency. The high margins of error means that these smaller communities may be subjected to either too much money or too little money. Since the government has limited funding, a county receiving too much money can be detrimental to other counties who may need it. This ties into a county receiving too little money, because since their margin of error is not accounted for, the misallocation of funds further spirals.



